Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-12-21 survey/survey.tex
Big push on Data II.

\documentclass{article}
\usepackage{amssymb,amsthm}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}

@@

\begin{definition}
  A \emph{commutative semiring} (sometimes called a \emph{commutative
    rig} (commutative ri\emph{n}g without negative elements) consists of a
  set $R$, two distinguished elements of $R$ named 0 and 1, and two
  binary operations~$+$ and $\cdot$, satisfying the following
  relations for any $a,b,c \in R$:
\[\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}\]
\end{definition}

@@

\begin{definition}[homotopy]
\label{def:homotopy}
Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
$\forall x:A. f(x) = g(x)$. We denote this $f \sim g$.
\end{definition}

\noindent It is easy to prove that homotopies (for any given function
space $A \rightarrow B$) are an equivalence relation.

\begin{definition}[Quasi-inverse]
\label{def:quasi}
For a function $f : A \rightarrow B$, a \emph{quasi-inverse} is a
triple $(g, \alpha, \beta)$, consisting of a function
$g : B \rightarrow A$ and two homotopies
$\alpha : f \circ g \sim \mathrm{id}_B$ and
$\beta : g \circ f \sim \mathrm{id}_A$.
\end{definition}

\begin{definition}[Equivalence of types]
  Two types $A$ and $B$ are equivalent $A ≃ B$ if there exists a
  function $f : A \rightarrow B$ together with a quasi-inverse for $f$.
\end{definition}

\jc{talk about the definition without quasi-inverse? Continue with
types being a commutative semiring in the above sense.}

>>>>

\documentclass{article}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}

@@

\begin{defn}
  A \emph{commutative semiring} (sometimes called a \emph{commutative
    rig} (commutative ri\emph{n}g without negative elements) consists of a
  set $R$, two distinguished elements of $R$ named 0 and 1, and two
  binary operations~$+$ and $\cdot$, satisfying the following
  relations for any $a,b,c \in R$:
\[\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}\]
\end{defn}

@@

\begin{defn}[homotopy]
\label{def:homotopy}
Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
$\forall x:A. f(x) = g(x)$. We denote this $f \sim g$.
\end{defn}

\noindent It is easy to prove that homotopies (for any given function
space $A \rightarrow B$) are an equivalence relation.  The simplest
definition of the data which makes up an equivalence is then

\begin{defn}[Quasi-inverse]
\label{def:quasi}
For a function $f : A \rightarrow B$, a \emph{quasi-inverse} is a
triple $(g, \alpha, \beta)$, consisting of a function
$g : B \rightarrow A$ and two homotopies
$\alpha : f \circ g \sim \mathrm{id}_B$ and
$\beta : g \circ f \sim \mathrm{id}_A$.
\end{defn}

\begin{defn}[Equivalence of types]
  Two types $A$ and $B$ are equivalent $A \simeq B$ if there exists a
  function $f : A \rightarrow B$ together with a quasi-inverse for $f$.
\end{defn}

\noindent Why \emph{quasi}? The reasons are beyond our scope, but
the interested reader can read sections $2.4$ and $4$ in \cite{hott-book}.
There are several conceptually different, but equivalent, 
``better'' definitions.  We record just one here:

\begin{defn}[Bi-invertibility]
\label{def:biinv}
For a function $f : A \rightarrow B$, a \emph{bi-inverse} is a
pair of functions $g,h : B \rightarrow A$ and two homotopies
$\alpha : f \circ g \sim \mathrm{id}_B$ and
$\beta : h \circ f \sim \mathrm{id}_A$.
\end{defn}

\noindent We can then replace quasi-inverse with bi-invertibility in
the definition of type equivalence. The differences will not matter to
us here.

If we replace $=$ by $\simeq$, then we can indeed prove that
types (with $\bot, \top, \uplus, \times$) form a semiring.

The reader familiar with universal algebra should pause
and ponder a bit about what we have done. We have lifted 
\emph{equality} from being in the signature of the ambient logic
and instead put it in the signature of the algebraic structure
of interest. Then, and only then, are we in a solid position
to show that we have a concept (semiring) which is flexible 
enough for types to be an instance.

If we revisit the Curry-Howard correspondance, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
  \qquad A \ \text{is inhabited} \]
The above is the real \textit{essence} of the correspondance.

But mere inhabitation falls far short of our goals of being
able to smoothly deform from one type to another.

Let's thus analyze the crux of the ``problem''. In logic, we have
that $\land$ and $\lor$ are both \emph{idempotent}: this is the
property of any binary operation $*$ where
$\forall a. a * a = a$. And it should be clear that idempotent
operations are \emph{forgetful}: its input has two copies of $a$,
but its output, only one. On the type side, something more subtle
happens. Consider $\top \uplus \top$ versus $\top$; the first has
\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
while the second only one ($\texttt{tt}$). These cannot be put in
bijective correspondance. Even though the ``payload'' \texttt{tt}
is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
away information -- something we have expressly disallowed.
Yes, this should remind you of Maxwell's demon: even though the
data is the same, they are tagged differently, and this
information is indeed information, and must be preserved.
